---
title: "Portable Interactive Control Charts"
output:
  html_document:
    self_contained: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Workaround for crosstalk bug which needs 'filter_slider' to be first HTML element when working with dates
# Simply create a hidden/'dummy' filter to initialise correct includes
# https://github.com/rstudio/crosstalk/issues/65
crosstalk::filter_slider("dummy", "dummy", crosstalk::SharedData$new(data.frame(d = as.Date("01/01/1970"))), ~d)
```

## Introduction

A key limitation for interactive plots/charts in R is the need for a Shiny server in order to handle the data processing and statistical calculations. The `controlcharts` package overcomes this by performing all calculations in JavaScript, allowing the charts to be fully portable and interactive without the need for a server. By constructing charts using `crosstalk` data objects, multiple charts can also be linked together to allow for interactive filtering and highlighting in a fully portable format.

The `controlcharts` package also performs all necessary data aggregations and pre-processing in JavaScript. This means that users do not need to pre-aggregate or manipulate their data before creating charts, simplifying the workflow and making it easier to create interactive visualisations directly from raw data. This also allows for dynamic data exploration, as users can filter and interact with the charts without needing to re-run R code.

## Generate Data

Consider a dataset containing monthly numerators and denominators for 10 different organisations over a 2-year period:

```{r}
library(controlcharts)

set.seed(12324)

dat <- do.call(
  rbind.data.frame,
  lapply(toupper(letters[1:10]), function(grp) {
    denom <- sample(0:100, 24)
    data.frame(
      country = sample(c("C1", "C2", "C3"), 1, replace = TRUE),
      organisation = grp,
      month_start = seq(as.Date('2024-01-01'), length.out=24, by="month"),
      numerators = rbinom(n = 24, size = denom, prob = 0.3),
      denominators = denom
    )
  })
)
DT::datatable(dat)
```

## Basic Interactivity

SPC charts and funnel plots can be constructed from this data directly, with the resulting HTML plots supporting mouseover tooltips:

```{r}
spc_plt <- controlcharts::spc(data = dat,
                              keys = month_start,
                              numerators = numerators,
                              denominators = denominators,
                              outlier_settings = list(astronomical = TRUE,
                                                      shift = TRUE,
                                                      two_in_three = TRUE),
                              nhs_icon_settings = list(show_variation_icons = TRUE))

fun_plt <- controlcharts::funnel(data = dat,
                                 keys = organisation,
                                 numerators = numerators,
                                 denominators = denominators)


spc_plt$html_plot
fun_plt$html_plot
```

### Cross-plot Reactivity and Dynamic Calculations: Crosstalk

By converting the data to a `crosstalk::SharedData` object, multiple charts can be linked together to allow for interactive filtering and highlighting. As the aggregations, filtering, and statistical calculations are performed dynamically in JavaScript, these charts remain fully portable and interactive without the need for a Shiny server.

```{r}
# Convert data to crosstalk SharedData object
crosstalk_dat <- crosstalk::SharedData$new(dat)

# Create charts using crosstalk data
spc_plt <- controlcharts::spc(data = dat,
                              keys = month_start,
                              numerators = numerators,
                              denominators = denominators,
                              outlier_settings = list(astronomical = TRUE,
                                                      shift = TRUE,
                                                      two_in_three = TRUE),
                              nhs_icon_settings = list(show_variation_icons = TRUE))

fun_plt <- controlcharts::funnel(data = dat,
                                 keys = organisation,
                                 numerators = numerators,
                                 denominators = denominators)
```

We can then use `crosstalk` filters to allow users to dynamically filter the data displayed in both charts. As the filters are adjusted, the data aggregations and statistical calculations are updated in real-time within the charts:

```{r, eval=FALSE}
country_filter <- crosstalk::filter_select("countryFilter", "Country",
                                             crosstalk_dat, ~country)
org_filter <- crosstalk::filter_select("orgFilter", "Organisation",
                                         crosstalk_dat, ~organisation)
date_filter <- crosstalk::filter_slider("monthFilter", "Date",
                                        crosstalk_dat, ~month_start)
crosstalk::bscols(
  date_filter,
  org_filter,
  country_filter
)
crosstalk::bscols(
  fun_plt$html_plot,
  spc_plt$html_plot
)
```

```{r, echo=FALSE}
country_filter <- crosstalk::filter_select("countryFilter", "Country",
                                             crosstalk_dat, ~country)
org_filter <- crosstalk::filter_select("orgFilter", "Organisation",
                                         crosstalk_dat, ~organisation)
date_filter <- crosstalk::filter_slider("monthFilter", "Date",
                                        crosstalk_dat, ~month_start)
crosstalk::bscols(date_filter, country_filter, org_filter)
```

```{r, echo=FALSE}
crosstalk::bscols(
  fun_plt$html_plot,
  spc_plt$html_plot
)
```
